---
title: "Introduction OpenTree - Instruction file"
author: "DARTH workgroup"
date: "11/7/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Make sure you have a recent version from R
```{r}
getRversion()
```

You can check the newest version on the website from R
https://www.r-project.org

# Download OpenTree
OpenTree is currently only available via GitHub
To install a package from GitHub you can use install_github
With the repository name in it
```{r}
# help packages
if (!require('pacman')) install.packages('pacman'); library(pacman) # use this package to conveniently install other packages
# load (install if required) packages from CRAN
p_load("devtools", "scales", "ellipse", "lazyeval", "igraph",  "ggraph", "reshape2", "knitr", "stringr", "jsonlite", "rstudioapi", "tidyverse", "dampack")                                               

install_github("DARTH-git/OpenTree", force = TRUE) # (Un)comment if there is a newer version
```

# After you are 
```{r}
p_load_gh("DARTH-git/OpenTree")
```

## Create or open decision tree. 

The function `create_tree()` creates a blank tree and the function `open_tree()` opens an existing decision tree. 

*IMPORTANT*: since `create_tree()` always creates blank new tree, do not use it to access or modify an existing tree, or else the tree will get erased. Always use `open_tree()` to open and modify existing trees.

Any changes made to the tree in OpenTree are automatically saved as a `.json` file to the working directory. If you are running it in an R script, the `.json` file will be saved to the path on your machine specified in `dir_name`. If you are running it in an R markdown document, the `.json` file will be saved to the path where the R markdown document is located. 

```{r}
create_tree(file_name = "DemoTree", dir_name = getwd())
open_tree(file_name = "DemoTree", dir_name = getwd())
```



# Define parameter input values

```{r}
v_names_str    <- c("Intervention", "No intervention")  # names of strategies
n_str          <- length(v_names_str)                   # number of strategies
wtp            <- 100000                                # willingness to pay threshold

# Probabilities
p_S       <- 0.52   # probability of getting sick
p_S_int   <- 0.40    # probability of getting sick under intervention


# Costs
c_S       <- 1200   # cost of viral encephalitis care without complications
c_H       <- 9000   # cost of viral encephalitis care with complications

# LE
le_S      <- 7      # remaining life expectancy for sick people
le_H      <- 10     # remaining life expectancy for healthy people

```



Extract probability weights and outcomes of the decision tree.

```{r, warning = F}
# extract the probability weights and outcomes  
df_tree <- evaluate_model("DemoTree", n_payoffs = 2)
```





Compute total cost and QALYs.

```{r}
# vector of total cost and QALYs
v_total_qaly <- v_total_cost <- vector(mode = "numeric", length = n_str)

# Calculate total costs and QALYs for each strategy 
 for (i in 1:n_str) {
  v_total_qaly[i] <- df_tree[[i]]$prob %*% df_tree[[i]]$payoff1
  v_total_cost[i] <- df_tree[[i]]$prob %*% df_tree[[i]]$payoff2
 }

# calculate vector of nmb
v_nmb <- v_total_qaly * wtp - v_total_cost 
          
df_output <- data.frame(Strategy =  v_names_str,
                        Cost     =  v_total_cost,
                        Effect   =  v_total_qaly,
                        NMB      =  v_nmb)

# model output
df_output
```

## 05 Cost-Effectiveness Analysis

```{r}
# create the transition probability matrix for NO treatment
decision_tree_demo_cea  <- calculate_icers(cost       = df_output$Cost,
                                          effect      = df_output$Effect,
                                          strategies  = df_output$Strategy)
decision_tree_demo_cea
```

## 05.1 Plot frontier of Decision Tree

```{r}
plot(decision_tree_demo_cea, effect_units = "QALYs", label="all")
```

